Notes on slotted e-graphs

### 2025-07-04 POPL2026/SlottedNotes.tex
notes update

>>>>

\section{Explicit substitution}

Lambda calculus with explicit substitution (also $\lambda_{\sigma}$-calculus) made the substitution operator a first-class element of the calculus (rather than a meta-level operation) to bridge the gap between lambda calculus and its concrete implementations~\cite{Abadi_Cardelli_Curien_Lévy_1991}.
However, the existence of unique normal forms turned out to be a delicate matter and to achieve uniqueness, lambda terms must be quotiented by \textit{graphical} equivalence in addition to $\alpha$-equivalence~\cite{Standardisation}.
We claim that using e-hypergraphs over e-graphs (or slotted e-graphs) may be more beneficial for doing equational reasoning for lambda calculus with explicit substitution.

The rules can be seen in a table below (plust a standard rule for $(\lambda x . u)v = u[x / v]$)

\begin{tabular}{lcc}
\texttt{subst\_same\_var} & $x[x / u] = u$ &\\
\texttt{subst\_diff\_var} & $y[x / u] = y$ & if $x \not \in \mathcal{F}(u)$\\
\texttt{subst\_application} & $(uu')[x / v] = u[x / v]u'[x / v]$ & \\
\texttt{subst\_lambda} & $(\lambda x . u)[y / v] = \lambda x . (u[y / v])$ & if $x \not \in \mathcal{F}(v)$ \\
\texttt{subst\_interchange\_1} & $(u[x / v])[x' / v'] = (u[x' / v'])[x / v[x / v']]$ & if $x' \in \mathcal{F}(v)$ and $x \in \mathcal{F}(v')$\\
\texttt{subst\_interchange\_2} & $(u[x / v])[x' / v'] = (u[x / v[x' / v']])[x / v[x / v']]$ & if ???\\
\texttt{subst\_comm} & $(u[x / v])[x' / v'] = (u[x' / v'])[x / v]$ & if $x \not \in \mathcal{F}(v')$ and $x' \not \in \mathcal{F}(v)$\\
\end{tabular}

while such rules are absorbed when using string diagrams (or hypergraphs) akin to proof nets, they are not absorbed by e-graph representation.
Below one could see the number of e-nodes for different set of equations enabled for saturating an expression \texttt{plus 2 2} using Church encoding.

\begin{tabular}{lcccc}
  Set of equations & built-in subst & explicit subst & explicit subst & explicit subst\\
  &&& + interchange&  + interchange\\
  &&&&  + comm\\
  \# of e-nodes & 40 & 161 & 289 & 356
\end{tabular}

String diagrams represent the same equivalence without loss of generality and without handling equations for substitutions explicitly.
